### 2024-03-20 doc/15yrsChange.Rmd
Update author list
reknitting with figure and text updates

Jennifer Jordan^1,4^

@@

4. Department of _____, George Mason University, Fairfax, Virginia 22030, USA

@@

regen_failure = 6 #some figure illustrating absence of small canopy trees/ shifting  size distributions

@@

- white-tailed deer (*Odocoileus virginianus* ) is important endemic nuisance species, over-abundant because of human influence

@mcgarvey_effects_2013
@holm_interactive_2013
@knauer_understory_2023

@@

Here, we use 15 years of detailed census data from a large forest dynamics plot including a 4-ha deer exclosure to test the following hypotheses:
(1) ) Aboveground C stocks in living biomass have declined since the initial census (-∆AGB)
(2) canopy tree mortality and associated woody biomass loss are increasing, in large part due to nonindigenous nuisance species
(3) Recruitment and woody growth have not kept pace with tree mortality, suggesting that future mortality will result in substantial net biomass loss and sustained decrease in the C storage potential of this ecosystem.

@@

The study was conducted in the 25.6-hectare Large Forest Dynamics plot at the National Zoo and Conservation Biology Institute in Front Royal, Virginia (38 53’36.6“N, 78 08’43.4 “W).  The plot, which is located in the central Appalachian Mountains adjacent to Shenandoah National Park, is composed of mature secondary eastern deciduous forest. Situated in the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus spp.*), and hickory (*Carya spp.*) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*) and witch hazel (*Hamamelis virginiana*) (SCBI site description). The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 (SI archives, Bourg et al. 2013).  Natural disturbances at the plot consist of wind and infrequent ice storms (Anderson-Teixiera 2015).  The plot, which includes a four-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters. It is one of 78 sites in the Forest Global Earth Observatory (ForestGEO), a global network of forest dynamic plots that promotes comparative forest ecology studies around the world.

* Add information about the history of pests and pathogens at this site and at similar forests

In addition to our consideration of the plot as a whole, we give special consideration to three portions of the plot, all upland forest (i.e., excluding low-lying areas around streams):
(1) low deer, low non-endemic insects & pathogens - 4 ha portion of upland forest in the SE quarter of the plot, fenced in YEAR and maintained with only occasional deer presence for the past ## years and with relatively low abundance of canopy species affected by non-endemic pests and pathogens;
(2) high deer, low non-endemic insects & pathogens # ha of upland forest outside the deer exclosure with relatively low abundance of canopy species affected by non-endemic pests and pathogens; 
(3) high deer, high non-endemic insects & pathogens - a # ha section of upland forest outside the deer exclosure with relatively high abundance of canopy species affected by non-endemic pests and pathogens. 
These areas were delineated as follows.
Upland forest was defined according to topographic wetness index, originally calculated for the SCBI plot in @mcgregor_tree_2021.
Vulnerability to non-endemic insects & pathogens was defined based on the abundance of canopy species affected by non-endemic pests and pathogens at the time of plot establishment (2008).

![**Figure `r map_and_photos`. Photos within the SCBI ForestGEO plot: (a) low deer, low non-endemic insects & pathogens, (b) high deer, low non-endemic insects & pathogens, (c) high deer, high non-endemic insects & pathogens ** *(add a map of the plot)* All photos taken September 2023 by K. Anderson-Teixeira.](display/Figure1.png)

@@

A number of field surveys are regularly conducted at the site, two of which are included in our analysis.  As part of the ForestGEO network, the plot undergoes a comprehensive woody plant inventory every five years, according to the protocol outlined in Condit (1998).  This inventory, hereby referred to as the census, includes all stems greater than 1 centimeter in diameter breast height (1.3 meters; dbh).  The census records information regarding the dbh, species, status and location of each stem included in the survey.  Each invidual is assigned an identifying number for sequential data collection in subsequent censuses and outfitted with a metal tag in the field. In multi-stemmed individuals, each additional stem that surpasses the 1 cm dbh threshold receives a stem number and associated tag.  Location is recorded on a digitized map of the plot on the ESRI ArcGIS FieldMaps application (*citation*).  Established in 2008, there have been four censuses at the site, comprising 15 years of detailed forest dynamics data. In addition to the ForestGEO census, since 2014 an annual mortality survey has been conducted on the site. Initiated, in part, as a response to the presence of Emerald Ashborer, the mortality survey is conducted on all trees that had a dbh greater than 10 centimeters in the previous full census. Through assessing individual tree health, this study illustrates trends in forest mortality and identifies factors that are associated with death. Data is collected on the current status, canopy position, percentage of crown living and intact, and visible indicators of poor tree health, such as physical damage, potential pathogens and insect infestation.

To improve the quality of the data, GitHub Actions continuous integration was implemented for quality control (Kim et al., 2022). Data from the previous census was used to populate the Field Maps app, and fields were created to input data. There were checks built into the app to help catch errors, for example the DBH field would flag a suspiciously high number and would not allow the form to be submitted if it was blank. To use the app in the field the census crew downloaded offline areas of where they were working and synced them to an online map after collecting data. Next the data was downloaded from ArcGIS online and pushed to GitHub, where it triggered a GitHub Action running data quality assurance and control (QA/QC) scripts. The scripts were written in R and checked the data for errors such as “suspicious positive growth.” After going through the QA/QC process, a csv of errors and warnings with the associated stem were output which could then be corrected in the field.

## Analyses

* We  
*

@@

![**Figure `r biomass_trends`. Aboveground biomass decreased at the site for the first time since establishment. The net decrease at the plot-level was driven by large losses in regions with high vulnerability to nuisance species.**](display/Figure2.jpeg)

@@

![**Figure `r loss_gain_net`. Fluxes of aboveground woody biomass. Large increases in woody mortality combined with sustained decreases in woody productivity result in decreasing net biomass in the most recent census.**](display/Figure3.jpeg)

@@

![**Figure `r mortality_by_genera`. Annual woody mortality flux by species/genera. We report large increases in mortality of white ash, red oak, and black oak.**](display/Figure4.jpeg)

@@

![**Figure `r recruitment_by_genera`. Annual recruitment of canopy & understory species/genera across all censuses. Recruitment of canopy species is significantly higher within the deer exclosure. Recruitment of understory species is 20x that of canopy species.**](display/Figure5.jpeg)

@@

@holtmann_carbon_2021

@@

>>>>

Jennifer Jordan^1,4^ ( Orcid ID : 0000-0001-5521-360X)

@@

4. Department of Environmental Science and Public Policy, George Mason University, Fairfax, Virginia 22030, USA

@@

regen_failure = 6 #2-panel - map & barplot illustrating areas losing canopy species in the understory and biomass in the overstory

@@

@@

Understanding the combined effects of multiple nuisance species on forest ecosystem processes requires data on both the forest understory and canopy over extended time periods. Here, we use 15 years of detailed census data from a large forest dynamics plot including a 4-ha deer exclosure to test the following hypotheses: (1) Aboveground C stocks in living biomass have declined since the initial census (-∆AGB) (2) canopy tree mortality and associated woody biomass loss are increasing, in large part due to non-indigenous nuisance species (3) Recruitment and biomass gains from tree growth have not kept pace with tree mortality, suggesting that future mortality will result in substantial net biomass loss and sustained decrease in the C storage potential of this ecosystem.

@@

The study was conducted in the 25.6-hectare Large Forest Dynamics plot at the National Zoo and Conservation Biology Institute in Front Royal, Virginia (38$^\circ$ 53' 36.6''N, 78$^\circ$  08’ 43.4''W). The plot, which is located in the central Appalachian Mountains adjacent to Shenandoah National Park, is composed of mature secondary eastern deciduous forest. Situated in the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus* spp.), and hickory (*Carya* spp.) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*). The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 (SI archives, Bourg et al. 2013).  Natural disturbances at the plot consist of wind and infrequent ice storms (Anderson-Teixiera 2015).  The plot, which includes a four-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters. It is one of 78 sites in the Forest Global Earth Observatory (ForestGEO), a global network of forest dynamic plots that promotes comparative forest ecology studies around the world.

In addition to our consideration of the plot as a whole, we give special consideration to three portions of the plot, all upland forest (i.e., excluding low-lying areas around streams):
(1) Low deer, low canopy vulnerability: A 4-ha portion of upland forest in the SE quarter of the plot, fenced in 1990 and maintained with only occasional deer presence for the past 24 years that has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens (Holm *et al.* 2013);
(2) high deer, low canopy vulnerability: 16.3 ha of upland forest outside of the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens; 
(3) high deer, high canopy vulnerability: 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens. 
Upland forest was defined according to the topographic wetness index, originally calculated for the SCBI plot in @mcgregor_tree_2021.
Vulnerability to non-indigenous insects and pathogens was defined based on the abundance of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment. The following canopy species were classified as affected: ash (*Fraxinus* spp.), elm (*Ulmus* spp.), butternut (*Juglans cinerea*), and American chestnut (*Carya dentata*).

![**_Figure `r map_and_photos`. Map and photos of the Forest Global Earth Observatory (ForestGEO) plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA._** Our analysis focuses on three habitat types: (a) low deer density, low abundance of canopy trees vulnerable to non-indigenous insects & pathogens, (b) high deer density, low abundance of canopy trees vulnerable to non-indigenous insects & pathogens, and (c) high deer density, high abundance of canopy trees vulnerable to non-indigenous insects & pathogens. *All photos taken September 2023 by K. Anderson-Teixeira; aerial photography from ESRI World Imagery*.](display/Figure1.png)

@@

As part of the ForestGEO network, the plot undergoes a comprehensive woody plant inventory every five years according to the protocol detailed in Condit (1998). This inventory, hereafter referred to as the census, includes all stems greater than one centimeter in diameter at breast height (1.3 meters; dbh).  The census records information regarding the dbh, species, living status and location of each stem included in the survey.  Each individual is marked with an identifying number for sequential data collection in subsequent censuses. In multi-stemmed individuals, each additional stem that surpasses the dbh threshold receives a stem number and associated tag. The plot was established in 2008 and recensused three times since, comprising 15 years of forest dynamics data. In the last census, data cleaning was continuously implemented 1) in an ESRI ArcGIS Field Maps application via built-in checks and 2) on a daily basis, via error reports generated by GitHub actions (Kim et al., 2022). Each individual's location was recorded in the Field Maps application using a digitized map of the plot (*citation*). Additional checks and fixes were performed at the conclusion of the last census. In addition to the ForestGEO census, since 2014, a tree mortality survey has been conducted annually at the site. Initiated, in part, as a response to the presence of emerald ash borer (*Agrilus planipennis*), the mortality survey is conducted on all trees that had a dbh $\geq$ 10 centimeters in the previous census. Through assessing individual tree health, we document trends in forest mortality and identify factors that are associated with death. Data is collected on the current status, canopy position, percentage of crown living and intact, and visible indicators of poor tree health, such as physical damage, potential pathogens and insect infestation.

## Calculating biomass dynamics and stem recruitment

We estimated total aboveground biomass (AGB) from the measured dbh of each stem in each census using extra-tropical allometric equations from the R package *allodb* (Gonzalez-Akre, 2024). *allodb* creates species-specific allometric equations through weighting of all applicable allometries within its database filtered by tree size, geography, and taxa information. Using the calculated stem AGB, we estimated AGB stocks at each census year for each of the plot regions and the plot as a whole. We evaluated trends in AGB over time by calculating annualized AGB fluxes including: 1) the net change in AGB, 2) aboveground woody growth (AWG; the addition of woody biomass from tree growth), 3) aboveground woody mortality (AWM; the change of living biomass to necromass due to tree mortality), and 4) aboveground woody recruitment (AWR; the addition of living biomass through recruitment) for each census interval and plot region. We further assessed the role of nuisance species on mortality by calculating AWM flux for each year and each species using the annual mortality survey data.

In addition to the estimation of AWR biomass flux, we analyzed the effects of nuisance species on recruitment through counts of newly recruited stems. We categorized each species as either canopy or understory based on expert knowledge of life history traits. *Quercus* spp. and *Carya* spp. were grouped by genera to better assess differences between the recruitment of canopy and understory species. We then calculated the number of new stems for each census interval for each  species or genera. Finally, we assessed the spatial overlap of biomass loss and recruitment of canopy species. For each .4 hectare quadrat, we calculated the net change in AGB from 2008 to 2023 and AGB stock of canopy species in the understory in 2023.

@@

![**_Figure `r biomass_trends`. Aboveground biomass at the SCBI ForestGEO plot since 2008 establishment._** Values are shown for the plot as a whole and for habitats differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens.](display/Figure2.jpeg)

@@

![**_Figure `r loss_gain_net`. Changes in aboveground woody biomass at the SCBI ForestGEO plot over each census interval since 2008 establishment, including (a) net change, (b) losses to mortality, and gains from (c) woody productivity and (d) recruitment._** Values are shown for the plot as a whole and for habitats differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens.](display/Figure3.jpeg)

@@

![**_Figure `r mortality_by_genera`. Annual woody mortality by species at the SCBI ForestGEO plot from 2008-2023._** Shown are the seven canopy taxa with highest average mortality rate in descending order and a category including all others. The census included all stems ≥ 10cm in the most recent full census.](display/Figure4.jpeg)

@@

![**_Figure `r recruitment_by_genera`. Average annual recruitment (2008-2023) of canopy and understory tree species at the SCBI ForestGEO plot across habitats differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens._** Note that y-axes differ, with understory species recruitment >20x that of canopy species recruitment.](display/Figure5.jpeg)

@@

We found that the combination of decreasing biomass (driven by increasing woody mortality) and limited recruitment of canopy species in the understory is compromising the potential the carbon sink, now and into the future. 
Differences in AGB were the biggest in the regions of the plot with high vulnerability to nuisance species. These regions also had the lowest amount of canopy species in the understory. Areas with low deer population had the highest amount of canopy species in the understory.
Some of the plot is increasing biomass, notably those with low proportions of vulnerable species. On net, these increases are not outpacing mortality. 
Suggests that the combination of heavy deer browsing on canopy species may be exacerbated in areas with canopy openings due to pathogen-derived mortality where understory species that are resistant to deer and fast growers can take advantage, further competing with canopy species.
While Emerald Ash-borer has had a large impact on biomass loss at our site, mortality among oak species, which is not linked to a non-indigenous pest, is substantial as well. Oak species in the region are generally found in older, even aged cohorts, which contributes to accelerated mortality rates (Oak et al., 2016). However, large oak mortality events, which often occur over protracted periods of time, have been common globally over the past century. These mortality events are often termed “oak decline” events due to the complex abiotic and biotic factors which interact to incite mortality (Manion and Lachance, 1992). Oak decline is believed to be the result of an imbalance in carbohydrate dynamics, which weakens oak trees’ resistance to pest and pathogen invasion, eventually leading to mortality (Manion and Lachance, 1992).

Using current understory as a predictor of the future canopy is imperfect. 
Some canopy species (examples) recruit in pulse events - if there is a shift in deer population and enough survival that coincides with a mast year than you could get regeneration into the canopy.
However, current recruitment levels are below a rate at which we would expect in maturing secondary forests and may be insufficient to prevent large periods without a close canopy.
Our data is from just the single site, which can make generalizability challenging. However, there is a lot of support that the drivers we identify here are representative of the region.
Mixed hardwood forests are typical of mid-Atlantic forests and deer populations are high throughout the region.
Vulnerable species (ash) are also common throughout these forests, and emerald ash borer continues to spread. Where it does spread it consistently decimates adult ash.
Our results underscore the importance of accounting for nuisance species in higher-level models and understanding of temperate forests. 
Nuisance species (driven by human pressures/actions) interact with other anthropogenic changes to modify forest ecosystems. 
Have the capacity to alter short-term dynamics: large carbon losses that would not be expected in models that don’t exclude it.
Longer-term have capacity to alter successional dynamics: can delay recovery from disturbances (including other disturbances) or even suppress regeneration into a closed-canopy for sustained periods. 
Decrease resilience to other pressures as pests and pathogens decrease diversity through shrinking the number of possible canopy species.
Positive takeaways: this work highlights the potential of active management. Preserving the forests that we have is more valuable than planting new forests. 
Pests and pathogens are driving mortality - may be opportunities for treatment/prevention of spread. 
Research into natural immunity (ash species in original range of EAB have immunity).
Management of deer populations to decrease browsing pressure - deer exclosure shows the importance to maintain an understory.

@@

Expand analyses across other temperate forests. Different forest compositions and herbivory dynamics, but many pests and pathogens. Leverage other forest dynamics sites to quantify the impact of nuisance species across large regions.
Investigate productivity responses of surviving canopy. We would expect survivors adjacent to gaps increase productivity as competition decreases, but may be hampered by climate change.
Interactions of global change drivers are critical to understand (though difficult to study).
